---
title: "STAT448 Assignment 1"
author: "David Ewing"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
geometry: a4paper, top=2cm
fontsize: 12pt
header-includes:
  - \renewcommand{\thesection}{\arabic{section}}
  - \renewcommand{\thesubsection}{\arabic{section}.\alph{subsection}}
---

# Question 1 - Algebraic Method

Observed values :

\[
\begin{aligned}
(Y) & : \{3, 9, 15\} - \text{Response Variable }  \\
 (X) & : \{6, 7, 8\} - \text{Explanatory Variable }\\
\text{Model} & : Y = \beta_0 + \beta_1 X + \epsilon
\end{aligned}
\]

## Compute ordinary least squares estimates of the coefficients by hand

The least squares estimates of \( \beta_0 \) and \( \beta_1 \) are given by:

\[
\begin{aligned}
\hat{\beta_1} &= \frac{\sum (X_i - \bar{X}) (Y_i - \bar{Y})}{\sum (X_i - \bar{X})^2} \\
\hat{\beta_0} &= \bar{Y} - \hat{\beta_1} \bar{X}
\end{aligned}
\]

 
 Mean Calculation
```{r}
x <- c(6, 7, 8)
y <- c(3, 9, 15)
x_bar <- mean(x)
y_bar <- mean(y)
x_bar
y_bar
```

\newpage

 Compute \( \hat{\beta_1} \)
```{r}
numerator <- sum((x - x_bar) * (y - y_bar))
denominator <- sum((x - x_bar)^2)
beta1_hat <- numerator / denominator
beta1_hat
```

 Compute \( \hat{\beta_0} \)
```{r}
beta0_hat <- y_bar - beta1_hat * x_bar
beta0_hat
```

## Calculate by hand the estimates of the residuals

The residuals are given by:
\[
\hat{\epsilon}_i = Y_i - (\hat{\beta_0} + \hat{\beta_1} X_i)
\]

```{r}
fitted_values <- beta0_hat + beta1_hat * x
residuals <- y - fitted_values
residuals
```

## (d) Compute Coefficients using Matrix Algebra in R

We express the model in matrix form:
\[
Y = X\beta + \epsilon
\]
where
\[
X = \begin{bmatrix}1 & 6\\ 1 & 7\\ 1 & 8\end{bmatrix}, \quad Y = \begin{bmatrix}3\\ 9\\ 15\end{bmatrix}
\]

The OLS estimator is:
\[
\hat{\beta} = (X^T X)^{-1} X^T Y
\]

```{r}
X_matrix <- cbind(1, x)
Y_matrix <- matrix(y, ncol = 1)
beta_hat_matrix <- solve(t(X_matrix) %*% X_matrix) %*% t(X_matrix) %*% Y_matrix
beta_hat_matrix
```

## (e) Compute Coefficients using `lm()` in R

```{r}
lm_model <- lm(y ~ x)
summary(lm_model)
```
